A sample computation

### 2024-07-10 draft_systoles.tex
add the SL_n case for n prime

\DeclareMathOperator{\assign}{\coloneqq}        % Assingment operator
\newcommand{\R}{\mathbb{R}}        % Real
\newcommand{\Q}{\mathbb{Q}}        % rationals
\newcommand{\C}{\mathbb{C}}        % Complex
\renewcommand{\P}{\mathbb{P}}        % Projective
\newcommand{\A}{\mathbb{A}}        % Affine
\renewcommand{\O}{\mathcal{O}}        % Structure sheaf
\newcommand{\Id}{\mathbf{Id}}        % Identity
\newcommand{\SL}{\mathbf{SL}_3(\mathbb{Z})}        % SL_3(Z)
\newcommand{\GL}{\mathbf{GL}_3} % GL_3
\newcommand{\PGL}{\mathbf{GL}_3} % PGL_3
\newcommand{\Mat}{\mathbf{Mat}_3(\mathbb{C})}        % Mat_3( C )
\newcommand{\MatZ}{\mathbf{Mat}_3(\mathbb{Z})}        % Mat_3( ZZ )
\newcommand{\SLp}{\Gamma_p}        % a congruence subgroup
\newcommand{\phip}{\phi_p}        % the golden ratio mod(p)
\newcommand{\hence}{\Rightarrow}        % hence
\newcommand{\minpoly}{\mathrm{minpoly}}        % minimal polynomial

@@

>>>>

\DeclareMathOperator{\assign}{\coloneqq}        % Assingment operator
\newcommand{\R}{\mathbb{R}}        % Real
\newcommand{\Q}{\mathbb{Q}}        % rationals
\newcommand{\C}{\mathbb{C}}        % Complex
\renewcommand{\P}{\mathbb{P}}        % Projective
\newcommand{\A}{\mathbb{A}}        % Affine
\renewcommand{\O}{\mathcal{O}}        % Structure sheaf
\newcommand{\Id}{\mathbf{Id}}        % Identity
\newcommand{\SL}{\mathbf{SL}_3(\mathbb{Z})}        % SL_3(Z)
\newcommand{\GL}{\mathbf{GL}_3} % GL_3
\newcommand{\PGL}{\mathbf{GL}_3} % PGL_3
\newcommand{\Mat}{\mathbf{Mat}_3(\mathbb{C})}        % Mat_3( C )
\newcommand{\Matn}{\mathbf{Mat}_n(\mathbb{C})}        % Mat_n( C )
\newcommand{\MatZ}{\mathbf{Mat}_3(\mathbb{Z})}        % Mat_3( ZZ )
\newcommand{\MatnZ}{\mathbf{Mat}_n(\mathbb{Z})}        % Mat_n( ZZ )
\newcommand{\SLp}{\Gamma_p}        % a congruence subgroup
\newcommand{\phip}{\phi_p}        % the golden ratio mod(p)
\newcommand{\hence}{\Rightarrow}        % hence
\newcommand{\minpoly}{\mathrm{minpoly}}        % minimal polynomial

\newcommand{\SLn}{\mathbf{SL}_n(\mathbb{Z})}        % SL_n(Z)

@@

\section{From $\SL$ to $\SLn$}

Consider
\begin{itemize}
\item $p > 2$ a prime number;
\item $\SLn$ the $n$-th special linear group over $\mathbb{Z}$;
\item $\SLp$ the $p$th congruence subgroup of $\SLn$;
\item $Z_G(x)$ the centralizer of $x \in G$;
\item $ f(t) \eqdef t^n - (1 + pt) (1 + 2pt) \ldots (1 + (n - 1) p t)$;
\item If $C$ is a matrix, $\chi_C(\lambda) \assign \det(\lambda \Id - C)$ is the characteristics polynomial.
\item Let $A$ be the irreducible operator with $\chi_A(t) = f(t)$;

For example, when $n = 4$ $A$ takes a form
\begin{equation}
A = 
\begin{pmatrix}
0 & 0 & 0 & 1 \\
1 & 0 & 0 & 6 p \\
0 & 1 & 0 & 11 p^2 \\
0 & 0 & 1 & 6 p^3
\end{pmatrix}
\in {\mathbf{SL}_4(\mathbb{Z})};        % SL_3(Z)
\end{equation}
\item 
Define $\tilde A \eqdef \Id + p A$.

For example, when $n = 4$ $\tilde A$ takes a form
\begin{equation}
\tilde A = \Id + p A =
\begin{pmatrix}
1 & 0 & 0 & p \\
p & 1 & 0 & 6 p^2 \\
0 & p & 1 & 11 p^3 \\
0 & 0 & p & 1 + 6 p^4
\end{pmatrix}
\in \SLp.
\end{equation}
\end{itemize}

\section{Intro2}
In this note we establish the following results:
\begin{enumerate}
\item
The regularor of $\tilde A$
grows as $O (\ln^{n-1} (p)) $;

\item
If $n$ is prime,
the index of the centralizers
$ [Z_{\SLn}(\tilde A) : Z_{\SLp}(\tilde A)] < C$
is bounded by some constant $C$ and does not depend on $p$.
\end{enumerate}
This might be interesting.

There are more general estimates due to Duke \cite{duke2003extreme} (see Section 3 Proposition 1).

Let $K \eqdef \Q(\alpha)$ be the number field obtain by attaching the root of $f(t)$ to $\Q$.
\begin{Prop}
$K / \Q$ is a totally real number field of degree $n$.
\end{Prop}

\begin{proof}
First, $f(t)$ is irreducible as it has no rational roots: 
\begin{gather}
f(1) = 1 - (1 + p) (1 + 2p) \ldots (1 + (n - 1) p) \ne 0,\\
f(-1) = (-1)^n - (1 - p) (1 - 2p) \ldots (1 - (n - 1) p ) \ne 0.
\end{gather}

Let's compute the roots of $f(t)$ approximately and show they are real.

Notice that for $|t| \ll 1$ all but one roots of $f(t)$ are close to the roots of 
\[
(1 + pt) (1 + 2pt) \ldots (1 + (n - 1) p t)
,\] 
them being $p^{-1} , \frac{1}{2} p^{-1} , \ldots, \frac{1}{(n-1) } p^{-1} $.

As all but one roots are real, the remaining one is also real.
\end{proof}

The next goal is to tackle the unit group $\O_K^*$ of $K$.

The Dirichlet unit theorem implies that $\O_K^*$ is an abelian group of rank $n-1$ (there might be some finite torsion of roots of unity.  It is bounded universally in $p$).
% depending only on $n$
% as the degree of the roots of unity is the totient function which grows as $n \to \infty$.
A basis of $\O_K^*$ modulo roots of unity is called a system of fundamental units.
Finding a system of fundamental units in general is a complicated task.

Instead, we shall find a finite index subgroup $U \subset \O_K^*$. 
% \red{maybe order $\O \subset \O_K$ instead?}
Then we shall bound above the regulator of $K$ with the regulator of $U$.

Moreover, using a Silverman's lower bound on the regulator of $K$ we bound above the index $[\O_K^*: U] < C_1$ with $C_1$ independent on $p$.
% \red{maybe say uniformly in $p$ instead?}

Define a subgroup 
\[
	U = (1 + p \alpha)^{\mathbb{Z}} \ldots (1 + (n-1) p \alpha)^{\mathbb{Z}} \subset \O_K^*
.\] 
\begin{Prop}
$U \subset \O_K^*$ is a finite index subgroup.
The regulator $\Reg_U$ is bounded above by $C_2 \log(p)^{n-1}$ where $C_2$ is a constant independent on $p$.
\end{Prop}

\begin{proof}
We have
\begin{gather}
\alpha^n - (1 + p\alpha) (1 + 2p\alpha) \ldots (1 + (n - 1) p\alpha) + 1 = 1  ,
\\
(1 + p\alpha) (1 + 2p\alpha) \ldots (1 + (n - 1) p\alpha) = \alpha^n.
\end{gather}
It shows that $1 + p \alpha, \ldots, 1 + (n - 1) p \alpha$ are indeed units.

Let $\alpha_1 < \ldots < \alpha_{n-1} < 0 $ be the negative roots of $f$.
We shall compute them approximately.
\begin{Lem}
\[
\alpha_k = \frac{-1}{k} p^{-1} + \frac{(-1)^{k + 1} }{ k\cdot  k! (n - k + 1)!} p^{-n-1}  + O(p^{-2 n - 1})
.\] 
\end{Lem}
\begin{proof}
In the first approximation $\alpha_k = \frac{-1}{k} p^{-1} + o(p^{-1})$.
Let's compute the residue term more persisely. 
Let $\alpha_k = \frac{-1}{k} p^{-1} ( 1 + \epsilon )$.
Substituting to $f(t)$, we get
\begin{gather*}
(1 + p\alpha_k) (1 + 2p\alpha_k) \ldots (1 + (n - 1) p\alpha_k) = \alpha_k^n, \\
- (- k p\alpha_k - k) (- 2 k p\alpha_k - k)  \ldots (- (n - 1) k p\alpha_k - k)  = (- k \alpha_k) ^n, \\
- (1 - \epsilon - k) (2 - 2 \epsilon - k)  \ldots ( (n-1) - (n-1) \epsilon - k)  = p^{-n} (1 + \epsilon)^n, \\
- (1 - k) (2 - k) \ldots (- 1) k \epsilon 1 \ldots ( (n-1) - k) + O ( \epsilon^2)
= p^{-n} + O( \epsilon p^{-n} ) , \\
(-1)^k k! ( n - k - 1)! \epsilon + O ( \epsilon^2)
= p^{-n} + O( \epsilon p^{-n} ) , \\
\epsilon = \frac{ (-1)^k p^{-n} }{ k! ( n - k - 1)! }  + O( \epsilon p^{-n} ) + O ( \epsilon^2), \\
\epsilon = \frac{ (-1)^k p^{-n} }{ k! ( n - k - 1)! }  + O( p^{-2n} ).
\end{gather*}
\end{proof}

By the definition of the regulator we have
\begin{multline}
\Reg_U = 
\left|
\begin{array}{cccc}
\ln |1 + p \alpha_1| 	& \ln |1 + p \alpha_2| & \ldots & \ln |1 + p \alpha_{n-1}| \\
\ln |1 + 2 p \alpha_1| 	& \ln |1 + 2 p \alpha_2| & \ldots &  \ln |1 + 2 p \alpha_{n-1}| \\
\vdots 	& \vdots & \ddots &  \vdots \\
\ln |1 + (n-1) p \alpha_1| 	& \ln |1 + (n-1) p \alpha_2| & \ldots &  \ln |1 + (n-1) p \alpha_{n-1}| 
\end{array} 
\right|
\\
\approx
\left|
\begin{array}{cccc}
\ln \left( \frac{ p^{-n}}{n!}  \right)  	& \ln \left( 1 - \frac{1}{2} \right)  & \ldots &  \ln \left( 1 - \frac{1}{n - 1} \right)  \\
\ln |1 - 2| 	& \ln \left( \frac{p^{-n} }{2\cdot  2! (n-1)!} \right)  & \ldots &  \ln \left( 1 - \frac{2}{n-1} \right)  \\
\vdots 	& \vdots & \ddots &  \vdots \\
\ln |1 - (n-1) | 	& \ln |1 - \frac{n-1}2| & \ldots &  \ln \left( \frac{p^{-n}}{(n-1)\cdot (n-1)! 2!} \right) 
\end{array} 
\right| \\
\approx n^{n-1} \log(p)^{n-1} 
.\end{multline}
\end{proof}
From the proof of the Proposition above one finds that the regulator of $U$ $ \Reg_U > 0 $ is positive for big enough prime $p$.
Hence the generators $(1 + p \alpha), \ldots, (1 + (n-1) p)$ of $U$ are multiplicatively independent, and $U \subset \O_K^*$ is a finite index subgroup.

We shall bound the index $[\O_K^* : U]$ from above via the following
\begin{Thm}[Silverman, \cite{silverman1984inequality}]
Let $K$ be a number field of degree $d$ with regulator $R$, and absolute discriminant $D_K$. 
Let $r(K)$ be the rank of the unit group in $K$, and let $p(K)$ be the maximum of $r(k)$ as $k$ ranges over proper subfields of $K$. 
We prove 
\[
R_K > c_d (\log \gamma_d  D_K )^{r(K) - p(K)} 
\] 
for constants $c_d, \gamma_d$ depending only on $d$. 
\end{Thm}
\red{It seems it's actually due to 
Robert Remak in
"Über Größenbeziehungen zwischen Diskriminante
und Regulator eines algebraischen Zahlkörpers" }

\begin{Prop}
Let $n$ be a prime.
Then the index $[\O_K^* : U] < C_3$ is bounded from above by a constant $C_3$ independent on $p$.
\end{Prop}
\begin{proof}
The discriminant of $K$ is the discriminant of the polynomial $f(t)$, which is approximately
\[
	\disc_K = \disc_f = O( p^{2 (n - 1)^2} p^{-(n - 1) (n - 2)} ) = O ( p^{n (n - 1) } )
.\] 
More precisely,
\[
	\disc_K = \disc_f \approx 
	1!^2 \ldots (n-1)!^2 p^{n (n - 1) }
.\]

Given that $n$ is prime and so there are no nontrivial subfield of $K$ the Silverman's theorem implies 
$$\Reg_K > C_4 (\log \disc_K)^{n-1} \approx n^{n-1} (n-1)^{n-1} C_4 \log(p)^{n-1}  > C_5\log(p)^{n-1}.$$

It is easy to see that 
\[
\Reg_U = [\O_K^* : U] \Reg_K
.\] 
Thus we have
\[
 C_2 \log(p)^{n-1} > \Reg_U = 
[\O_K^* : U] \Reg_K >
[\O_K^* : U] C_5 \log(p)^{n-1}
.\] 
It follows that the index is bounded from above
\[
[\O_K^* : U] < \frac{C_2}{C_5}
,\] 
as desired.
\end{proof}

\section{Centralizers}
\begin{Prop}
Let $n$ be a prime.  
Then the index of the centralizers
\[
	[Z_{\SLn}(\tilde A) : Z_{\SLp}(\tilde A)] < C_3
\] 
is bounded from above by a constant independent of $p$.
\end{Prop}
\begin{proof}
	We shall exploit the properties of the previously introduced number field $K = \mathbb{Q}(\alpha)$, where $\alpha$ is a root of $f(t) = t^n - (1 + pt) \ldots (1 + (n-1) pt)$.

Since $A$ is regular, 
% over $\mathbb{Z}$
there is an isomorphism of $\mathbb{Z}$-algebras

\[
\Phi: \mathbb{Z}[A] 
\mathop{\longrightarrow}^{\sim}
\mathbb{Z}[\alpha], A \mapsto \alpha
.\]

It identifies the stabilizer of $\tilde A$ in $\SLn$ with the multiplicative group of the order $\mathbb{Z}[\alpha] \subset \O_K$.
\[
Z_{\SLn}(\tilde A) 
= Z_{\SLn}(A) 
= \mathbb{Z}[A]^*
\mathop{\simeq }^{\Phi} 
\mathbb{Z}[\alpha]^*
.\]

The isomorphism $\Phi$ identifies
the subgroup $U \subset \mathbb{Z}[\alpha]^*$ with
$$
\Phi^{-1} (U) = \tilde A^{\mathbb{Z}} (2 \tilde A - \Id )^{\mathbb{Z}} \ldots   ((n-1) \tilde A - (n-2) \Id )^{\mathbb{Z}} \subset Z_{\SLp}(\tilde A) 
.$$

Thus, there is a chain of inclusions
\[
\Phi^{-1} (U) \subset Z_{\SLp}(\tilde A) 
\subset Z_{\SLn}(\tilde A) = \Phi^{-1} \mathbb{Z}[\alpha]^*
.\] 
Therefore, there is a bound
\[
[Z_{\SLn}(\tilde A) : Z_{\SLp}(\tilde A) ]
\le
[\mathbb{Z}[\alpha]^*, U]
\le 
[\O_K^*, U] < C_3
.\] 
\end{proof}
